Characteristic polynomials of p-adic matrices

### 2017-04-16 charpoly/charpoly.tex
Switch comatrix to adjugate and adjust notation

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\PP}{PP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\rec}{rec}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\disc}{Disc}
\DeclareMathOperator{\row}{row}
\DeclareMathOperator{\col}{col}

\newcommand{\N}{\mathbb N}
\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\R}{\mathbb R}
\newcommand{\OK}{\mathcal{O}_K}

@@

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi(A)$ of
an $n \times n$ $p$-adic matrix $A$ using differential precision methods
developed previously.  When $A$ is integral with precision $O(p^N)$,
we give a criterion (checkable in time $\softO(n^\omega)$) for
$\chi(A)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
algorithm for determining the optimal precision when the criterion is not
satisfied, and give examples when the precision is larger than $O(p^N)$.
\end{abstract}

@@

      \xavier{One can argue that we do not need \emph{a priori} much accuracy on the precision of $\com(A-X)$ because only the valuations of the entries of the latter matrix are relevant. So counting operations in $K$ makes sense. By the way, we could say at some point that a good idea would be to use relaxed algorithms and increase the precision until we are able to read the valuations.}

@@

\item \tdo{The definition of comatrix is usually the matrix formed by the cofactors such that $A^{-1} = \com(A)^T / \det(A)$.}
      \xavier{Ok pour switching $\com$ to $\text{Adj}$ (and adding a sentence explaining the difference with the previous paper).}

@@

\item \done{We have switched from $A^t$ to $A^T$ to denote the transpose.}

@@

\item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing com(A) differ), would be welcome.}

@@

\item \tdo{Define comatrix properly, as requested by reviewer 3.}

@@

Thanks to the application of the framework of differential precision
of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
we know that the precision of the characteristic polynomial
$\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
is determined by the comatrix $\com(X -M).$
In this article, we provide:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item Proposition \ref{prop:shortcom}: a factorization of $\com(X -M)$
as a product of two rank-$1$ matrices (when $M$ has a cyclic vector),
computable in $\softO(n^\omega)$ operations by Theorem \ref{thm:compute_shortcom}
\item Corollary \ref{cor:prec_gain}: a simple, $\softO(n^\omega)$ criterion
to decide whether $\chi_M$ is defined at precision
higher than the precision of $M$ (when $M \in M_n(\mathbb{Z}_p)$).
\item Theorem \ref{thm:alg3}: a $\softO(n^3)$ algorithm with operations in $\Zp$
to compute the optimal precision on each coefficient of $\chi_M$ (when $M$
is given with uniform precision on its entries).
\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
algorithm to compute the optimal precision on each eigenvalue of $M$
\end{enumerate}

@@

In Section \ref{sec:theo_study}, we review the differential
theory of precision developed in \cite{caruso-roe-vaccon:14a}
and apply it to the specific case of the characteristic polynomial,
giving conditions under which the differential will be surjective
(and thus provide a good measure of precision).  We also give
a condition based on reduction modulo $p$ that determines
whether the characteristic polynomial will have a higher
precision than the input matrix, and show that the image
of the set of integral matrices has the structure of an $\OK[X]$-module
when $M$ is itself integral. Finally, we give a compact description
of $\com(X-M)$, the main ingredient in the differential.

In Section \ref{sec:diffHess}, we develop $\softO(n^3)$ algorithms to approximate
the Hessenberg form of $M$, and through it to find $\com(X-M)$ and thus
find the precision of the characteristic polynomial of $M$.  In Section \ref{sec:diffFrob},
we give a $\softO(n^\omega)$ algorithm to compute the compact description of $\com(X-M)$.

@@

\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. 
We will write that $f(n)=\softO (g(n))$ if there exists some
$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k).$
We will write $M$ for an $n \times n$ matrix over $K$,
and $\chi$ the characteristic polynomial map, $\chi_M \in K[X]$ for
the characteristic polynomial of $M$ and $d\chi_M$ for the differential
of $\chi$ at $M$, as a linear map from $M_n(K)$ to the space
of polynomials of degree less than $n$.
We fix an $\omega \in \mathbb{R}$ such that the multiplication
of two matrices over a ring is in $O(n^\omega)$ operations in the
ring. Currently, the smallest known $\omega$ is less than
$2.3728639 $ thanks to \cite{legall:14a}. 
We will denote by $I_n$ the identity matrix of rank $n$ in $M_n(K).$
When there is no ambiguity, we will drop this $I_n$
for scalar matrices, \textit{e.g.} for $\lambda \in K$
and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n - M.$
Finally, we write $\sigma_1(M), \dots, \sigma_n(M)$ for the
elementary divisors of $M$, sorted in increasing order of valuation.

@@

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$, we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi_M$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}

@@

For any $A \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
as the image of $d\chi$ at $AMA^{-1}$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the bottom row of
$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.

Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
where $\mu, d_\mu$ ranges over the eigenvalues and dimensions of the other Jordan blocks.
Since $\lambda$ occurs in two Jordan blocks,
$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi_M$ will not be surjective.
\end{proof}

@@

Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.

@@

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
multiplication by $A$.
\end{prop}
\begin{proof}
Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $\Lambda$ is given by
the $\OK$-span of the entries of $C$.   Using the fact that the product of a matrix
with its comatrix is the determinant, $(X - M) \cdot C = \chi_M$ and thus
$P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
of the left hand side is precisely $P(X) \cdot \Lambda$, while the span of the entries
of the right hand side is contained within $\Lambda$ since $P(M) \in M_n(\OK)$.
\end{proof}

@@

\begin{prop}
\label{prop:shortcom}
We keep the previous notations and assumptions.
Let $Y$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
\begin{equation}
\label{eq:shortcom}
\com(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T
\mod \chi_M
\end{equation}
for some $\alpha \in K[X]$.
\end{prop}

\begin{proof}
Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}\mathscr{C}$
modulo $\chi_M$. On the other hand, a direct computation shows that
every column vector $W$ lying in the right kernel of $X{-}\mathscr{C}$ modulo 
$\chi_M$ can be written as $W = w \cdot Y^T$ for some $w \in 
K[X]/\chi_M$. We deduce that $C \equiv P \cdot Y^T B \pmod{\chi_M}$
for some row vector $B$.
Applying the same reasoning with $M^T$, we find that $B$ can be
written $B = \alpha Y Q^T$ for some $\alpha \in K[X]/\chi_M$ and
we are done.
\end{proof}

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ entries to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Section \ref{sec:diffFrob} that interesting
information can be read off of this short form $(\alpha, P, Q, 
\chi_M)$.

\begin{rem}
With the previous notation, if $U \in GL_n(K),$
the quadruple for 
$UMU^{-1}$ is
$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
which can be computed in $O(n^\omega)$ operations in $K.$
This is faster than computing $U \com(X-M) U^{-1},$
which is, at first sight, in $O(n^4)$ operations in $K.$
\end{rem}

@@

In this section, we combine the computation of a Hessenberg form
of a matrix and the computation of the inverse through the Smith normal form (SNF)
over a complete discrete valuation field (CDVF)
to compute $\com(X-M)$ and $d \chi$.
If $M \in M_n(\OK)$, then only division by invertible
elements of $\OK$ will occur.
%The computation of a Hessenberg form provides an $O(n^3)$
%algorithm to compute the characteristic polynomial.

@@

\begin{prop} 
Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
\end{prop}
\begin{proof}
Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
with the $\varepsilon$'s being units,
the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\mathbb{F}_q[X],$ $\mathbb{Q}[X]$ or adequate extension) of $\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}.$
Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
The rest is clear.
\end{proof}

@@

\subsection{The comatrix of $X{-}H$}

In this section, we compute $\com(X-H)$ for a Hessenberg matrix $H$
using the Smith normal form computation of the previous section.
The entries of $\com(X-H)$ lie in $K[X]$, which is not a CDVF,
so we may not directly apply the methods of the previous section.
However, we may relate $\com(X-H)$ to $\com(1-XH)$, whose
entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\com(X-H)$
using an SNF method, with no division in $K$.

@@

\begin{lem} \label{lem:comatrix_of_similar}
If $M_1,M_2 \in M_n(K)$ and $P \in GL_n (K)$ are such that
$M_1=PM_2P^{-1},$ then:
\[ \com (X-M_1)=P \com (X-M_2) P^{-1}. \] 
\end{lem}

The second ingredient we need is reciprocal polynomials.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
We define the reciprocal polynomial of order $d$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
We denote by $A^{\rec,d}$ the matrix with $(A^{\rec,d})_{i,j} = (A_{i,j})^{\rec,d}$.
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
\com(1-XM)^{\rec,n-1}=&\com(X-M), \\
(\chi_M I_n)^{\rec,n}=&(1-XM) \com(1-XM).\\
\end{eqnarray*}
\end{lem}
\begin{proof}
It all comes down to the following result:
let $A \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
then $\det (A^{\rec,1})=\det(A)^{\rec,d}.$
Indeed, one can use multilinearity of the determinant on $X^d \det(A(1/X))$
to prove this result.
It directly implies the second part of the lemma; the first part follows
from the fact that the entries of $\com(X-M)$ and of $\com(1-XM)$
are determinants of size $n-1$.
\end{proof}

This lemma allows us to compute $\com(1-XM)$ instead of $\com(X-M).$
This has a remarkable advantage: the pivots during the computation of
the SNF of $\com(1-XM)$ are units of $\OK\llb X\rrb,$ and are known
in advance to be on the diagonal. This leads to a very smooth
precision and complexity behaviour when the input matrix lives in 
$M_n(\OK).$

@@

\noindent {\bf Algorithm 2:} {\tt Approximate $\com (X -H)$ }

@@

\begin{theo}
Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix.
Then, using Algorithm 2, one can compute $\com (X -H)$
and $\chi_H$ in 
$\softO (n^3)$ operations in $\OK$ at the precision given by $H.$
\end{theo}
\begin{proof}
First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
in $\OK$ at the precision given by $H.$
Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
the only other nonzero coefficient in its column is $U_{i+1,i}$.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $\OK \llb X \rrb$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q$ is done by operations on rows, starting from the identity matrix.
The order in which the entries of $U$ are cleared implies that $Q$ is just filled in as an upper triangular matrix:
no additional operations in $\OK\llb X\rrb$ are required. Thus the total cost
for lines 2 and 3 is indeed $\softO (n^3)$ operations.

@@

\subsection{The comatrix of $X{-}M$}

In this section, we combine Proposition \ref{prop:shortcom}
with Algorithm 2 to compute the comatrix of $X-M$
when $\chi_M$ is squarefree.   Note that this condition on
$\chi_M$ is equivalent to $M$ being diagonalizable under
the assumption that $d\chi_M$ is surjective.  The result is
the following $\softO(n^3)$ algorithm, where the only
divisions are for gcd and modular inverse computations.

@@

\noindent {\bf Algorithm 3:} {\tt Approximate $\com (X{-}M)$ }

>>>>

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\PP}{PP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\adj}{Adj}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\rec}{rec}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\disc}{Disc}
\DeclareMathOperator{\row}{row}
\DeclareMathOperator{\col}{col}

\newcommand{\N}{\mathbb N}
\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\Fq}{\mathbb{F}_q}
\newcommand{\R}{\mathbb R}
\newcommand{\OK}{\mathcal{O}_K}

@@

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi(M)$ of
an $n \times n$ $p$-adic matrix $M$ using differential precision methods
developed previously.  When $M$ is integral with precision $O(p^N)$,
we give a criterion (checkable in time $\softO(n^\omega)$) for
$\chi(M)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
algorithm for determining the optimal precision when the criterion is not
satisfied, and give examples when the precision is larger than $O(p^N)$.
\end{abstract}

@@

      \xavier{One can argue that we do not need \emph{a priori} much accuracy on the precision of $\adj(M-X)$ because only the valuations of the entries of the latter matrix are relevant. So counting operations in $K$ makes sense. By the way, we could say at some point that a good idea would be to use relaxed algorithms and increase the precision until we are able to read the valuations.}

@@

\item \done{We have switched to using the term ``adjugate'' rather than ``comatrix.''}

@@

\item \done{We have switched from $M^t$ to $M^T$ to denote the transpose.}

@@

\item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing adj(M) differ), would be welcome.}

@@

\item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}

@@

Thanks to the application of the framework of differential precision
of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
we know that the precision of the characteristic polynomial
$\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
is determined by the adjugate $\adj(X -M).$
In this article, we provide:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item Proposition \ref{prop:shortcom}: a factorization of $\adj(X -M)$
as a product of two rank-$1$ matrices (when $M$ has a cyclic vector),
computable in $\softO(n^\omega)$ operations by Theorem \ref{thm:compute_shortcom}
\item Corollary \ref{cor:prec_gain}: a simple, $\softO(n^\omega)$ criterion
to decide whether $\chi_M$ is defined at precision
higher than the precision of $M$ (when $M \in M_n(\mathbb{Z}_p)$).
\item Theorem \ref{thm:alg3}: a $\softO(n^3)$ algorithm with operations in $\Zp$
to compute the optimal precision on each coefficient of $\chi_M$ (when $M$
is given with uniform precision on its entries).
\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
algorithm to compute the optimal precision on each eigenvalue of $M$
\end{enumerate}

@@

In Section \ref{sec:theo_study}, we review the differential
theory of precision developed in \cite{caruso-roe-vaccon:14a}
and apply it to the specific case of the characteristic polynomial,
giving conditions under which the differential will be surjective
(and thus provide a good measure of precision).  We also give
a condition based on reduction modulo $p$ that determines
whether the characteristic polynomial will have a higher
precision than the input matrix, and show that the image
of the set of integral matrices has the structure of an $\OK[X]$-module
when $M$ is itself integral. Finally, we give a compact description
of $\adj(X-M)$, the main ingredient in the differential.

In Section \ref{sec:diffHess}, we develop $\softO(n^3)$ algorithms to approximate
the Hessenberg form of $M$, and through it to find $\adj(X-M)$ and thus
find the precision of the characteristic polynomial of $M$.  In Section \ref{sec:diffFrob},
we give a $\softO(n^\omega)$ algorithm to compute the compact description of $\adj(X-M)$.

@@

\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. 
We will write that $f(n)=\softO (g(n))$ if there exists some
$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k).$
We will write $M$ for an $n \times n$ matrix over $K$,
and $\chi$ the characteristic polynomial map, $\chi_M \in K[X]$ for
the characteristic polynomial of $M$ and $d\chi_M$ for the differential
of $\chi$ at $M$, as a linear map from $M_n(K)$ to the space
of polynomials of degree less than $n$.
We fix an $\omega \in \R$ such that the multiplication
of two matrices over a ring is in $O(n^\omega)$ operations in the
ring. Currently, the smallest known $\omega$ is less than
$2.3728639 $ thanks to \cite{legall:14a}. 
We will denote by $I_n$ the identity matrix of rank $n$ in $M_n(K).$
When there is no ambiguity, we will drop this $I_n$
for scalar matrices, \textit{e.g.} for $\lambda \in K$
and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n - M.$
Finally, we write $\sigma_1(M), \dots, \sigma_n(M)$ for the
elementary divisors of $M$, sorted in increasing order of valuation.

@@

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$, we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi_M$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
we have $\adj(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
d\chi_M: dM \mapsto \tr(\adj(X-M) \cdot dM).
\end{equation}

@@

For any $P \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
as the image of $d\chi$ at $PMP^{-1}$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the bottom row of
$\adj(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.

Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X-M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
where $\mu, d_\mu$ ranges over the eigenvalues and dimensions of the other Jordan blocks.
Since $\lambda$ occurs in two Jordan blocks,
$X - \lambda$ will divide every entry of $\adj(X-M)$ and $d\chi_M$ will not be surjective.
\end{proof}

@@

Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.

@@

\begin{prop}
Let $R$ be the subring of $K[X]$ consisting of polynomials $f$ for
which $f(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
multiplication by $R$.
\end{prop}
\begin{proof}
Let $A = \adj(X-M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
the $\OK$-span of the entries of $A$.   Using the fact that the product of a matrix
with its adjugate is the determinant, $(X - M) \cdot A = \chi_M$ and thus
$f(X) \cdot A \equiv f(M) \cdot A \pmod{\chi_M(X)}$.  The span of the entries
of the left hand side is precisely $f(X) \cdot \Lambda$, while the span of the entries
of the right hand side is contained within $\Lambda$ since $f(M) \in M_n(\OK)$.
\end{proof}

@@

\begin{prop}
\label{prop:shortcom}
We keep the previous notations and assumptions.
Let $Y$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
\begin{equation}
\label{eq:shortcom}
\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T
\mod \chi_M
\end{equation}
for some $\alpha \in K[X]$.
\end{prop}

\begin{proof}
Write $A = \adj(X{-}M)$. From $(X{-}M) \cdot A \equiv 0 
\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} A \equiv 0 \pmod{\chi_M}$. 
Therefore each column of $P^{-1} A$ lies in the right kernel of $X{-}\mathscr{C}$
modulo $\chi_M$. On the other hand, a direct computation shows that
every column vector $W$ lying in the right kernel of $X{-}\mathscr{C}$ modulo 
$\chi_M$ can be written as $W = w \cdot Y^T$ for some $w \in 
K[X]/\chi_M$. We deduce that $A \equiv P \cdot Y^T B \pmod{\chi_M}$
for some row vector $B$.
Applying the same reasoning with $M^T$, we find that $B$ can be
written $B = \alpha Y Q^T$ for some $\alpha \in K[X]/\chi_M$ and
we are done.
\end{proof}

Proposition~\ref{prop:shortcom} shows that $\adj(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ entries to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Section \ref{sec:diffFrob} that interesting
information can be read off of this short form $(\alpha, P, Q, 
\chi_M)$.

\begin{rem}
With the previous notation, if $U \in GL_n(K),$
the quadruple for 
$UMU^{-1}$ is
$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
which can be computed in $O(n^\omega)$ operations in $K.$
This is faster than computing $U \adj(X-M) U^{-1},$
which is, at first sight, in $O(n^4)$ operations in $K.$
\end{rem}

@@

In this section, we combine the computation of a Hessenberg form
of a matrix and the computation of the inverse through the Smith normal form (SNF)
over a complete discrete valuation field (CDVF)
to compute $\adj(X-M)$ and $d \chi$.
If $M \in M_n(\OK)$, then only division by invertible
elements of $\OK$ will occur.
%The computation of a Hessenberg form provides an $O(n^3)$
%algorithm to compute the characteristic polynomial.

@@

\begin{prop} 
Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
\end{prop}
\begin{proof}
Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
with the $\varepsilon$'s being units,
the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\Fq[X],$ $\Q[X]$ or adequate extension) of $\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}.$
Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
The rest is clear.
\end{proof}

@@

\subsection{The adjugate of $X{-}H$}

In this section, we compute $\adj(X-H)$ for a Hessenberg matrix $H$
using the Smith normal form computation of the previous section.
The entries of $\adj(X-H)$ lie in $K[X]$, which is not a CDVF,
so we may not directly apply the methods of the previous section.
However, we may relate $\adj(X-H)$ to $\adj(1-XH)$, whose
entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
using an SNF method, with no division in $K$.

@@

\begin{lem} \label{lem:adjugate_of_similar}
If $M_1,M_2 \in M_n(K)$ and $P \in GL_n (K)$ are such that
$M_1=PM_2P^{-1},$ then:
\[ \adj (X-M_1)=P \adj (X-M_2) P^{-1}. \] 
\end{lem}

The second ingredient we need is reciprocal polynomials.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $d \in \mathbb{N}$ and $f \in K[X]$ of degree at most $d.$ 
We define the reciprocal polynomial of order $d$ of $f$ as $f^{\rec,d}=X^d f \left( 1/X \right).$
Let $F \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
We denote by $F^{\rec,d}$ the matrix with $(F^{\rec,d})_{i,j} = (F_{i,j})^{\rec,d}$.
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
\adj(1-XM)^{\rec,n-1}=&\adj(X-M), \\
(\chi_M I_n)^{\rec,n}=&(1-XM) \adj(1-XM).\\
\end{eqnarray*}
\end{lem}
\begin{proof}
It all comes down to the following result:
let $F \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
then $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
Indeed, one can use multilinearity of the determinant on $X^d \det(F(1/X))$
to prove this result.
It directly implies the second part of the lemma; the first part follows
from the fact that the entries of $\adj(X-M)$ and of $\adj(1-XM)$
are determinants of size $n-1$.
\end{proof}

This lemma allows us to compute $\adj(1-XM)$ instead of $\adj(X-M).$
This has a remarkable advantage: the pivots during the computation of
the SNF of $\adj(1-XM)$ are units of $\OK\llb X\rrb,$ and are known
in advance to be on the diagonal. This leads to a very smooth
precision and complexity behaviour when the input matrix lives in 
$M_n(\OK).$

@@

\noindent {\bf Algorithm 2:} {\tt Approximate $\adj (X -H)$ }

@@

\begin{theo}
Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix.
Then, using Algorithm 2, one can compute $\adj (X -H)$
and $\chi_H$ in 
$\softO (n^3)$ operations in $\OK$ at the precision given by $H.$
\end{theo}
\begin{proof}
First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
in $\OK$ at the precision given by $H.$
Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
the only other nonzero coefficient in its column is $U_{i+1,i}$.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $\OK \llb X \rrb$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q$ is done by operations on rows, starting from the identity matrix.
The order in which the entries of $U$ are cleared implies that $Q$ is just filled in as an upper triangular matrix:
no additional operations in $\OK\llb X\rrb$ are required. Thus the total cost
for lines 2 and 3 is indeed $\softO (n^3)$ operations.

@@

\subsection{The adjugate of $X{-}M$}

In this section, we combine Proposition \ref{prop:shortcom}
with Algorithm 2 to compute the adjugate of $X-M$
when $\chi_M$ is squarefree.   Note that this condition on
$\chi_M$ is equivalent to $M$ being diagonalizable under
the assumption that $d\chi_M$ is surjective.  The result is
the following $\softO(n^3)$ algorithm, where the only
divisions are for gcd and modular inverse computations.

@@

\noindent {\bf Algorithm 3:} {\tt Approximate $\adj (X{-}M)$ }
